Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-02-08 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

\medskip
In immiscible displacements, viscous fingering occurs when the viscosity ratio is greater than unity. As surface tension becomes weak, the interface is stressed and becomes unstable leading to the formation of fingers. At this point it should be mentioned that there are two parameters -- Peclet number (Pe) and mobility ratio (MR), that determine the stability characteristics. There are always a few dominant fingers that spread and shield the growth of other fingers. The interface of the main finger collapses and starts splitting into new lobes of fingers. One of these new fingers may eventually outgrow the others and then spreads to occupy an increasingly larger width. In the process, the finger reaches a critical width while the saturation %concentration gradient 
of its front becomes steep as a result of stretching caused by the cross-flow, causing the tip of the finger to become unstable and splits again, and the pattern repeats itself. Therefore, surface tension plays an essential dual role, it must be weak enough for the tip front to be unstable, but it is also the physical force causing the spreading and ensuing repeated branching \citep{tan_1988, carvalho_2013}.

@@

\medskip
Fluid flow dynamics through heterogeneous porous media are sensibly more complex than in homogeneous media and strongly depend on problem properties. Spatial variation in geological formations occurs in all length-scales, where heterogeneity characteristics in small length-scales (\ie pore) are statistically embedded into permeability (absolute and relative) and porosity parameters. Such multi-scale heterogeneity induces preferential flow pathways and plays a significant role in the onset of fluid instabilities as it triggers fingers formation and their accelerated growth \citep[see][]{ewing_1989,tchelepi_1994}.

@@

%%%%%%%%%%%%%
\section{Conclusions}\label{Section:Conclusion}
% Short (1 paragraph) summary of the paper with focus on the main theoretical aspects that were proved by our work in the present manuscript;
The purpose of this paper is to numerically investigate the dynamics of viscous flow instabilities in porous media. Simulations involving formation and growth of viscous fingers were conducted with the CVFEM-based multi-fluid flow simulator Fluidity/IC-FERST model. The model is based upon a dual consistent pressure-velocity representation in CV and FEM spaces and the novel family of FE-pairs, \PN[n]{m} and \PNDG[n]{m}. Saturation and other scalar fields (e.g., density, temperature etc) are discretised in CV space and may be fully discontinuous across finite element boundaries. %Material absolute permeability is piecewise constant in FE space to ensure the correct representation of the surface.
In order to strongly enforce discretised Darcy equations at the boundaries between elements and control volumes, a directional-weighted flux-limited was introduced to take into account discontinuities risen from both, numerical formulation (based on discontinuous-Galerkin finite element method, DGFEM) and control volumes within finite elements.

>>>>

\medskip
In immiscible displacements, viscous fingering occurs when the viscosity ratio is greater than unity. As surface tension becomes weak, the interface is stressed and becomes unstable leading to the formation of fingers. At this point it should be mentioned that there are two parameters -- Peclet number (Pe) and mobility ratio (MR), that determine the flow stability characteristics. There are always a few dominant fingers that spread and shield the growth of other fingers. The interface of the main finger collapses and starts splitting into new lobes of fingers. One of these new fingers may eventually outgrow the others and then spreads to occupy an increasingly larger width. In the process, the finger reaches a critical width while the saturation %concentration gradient 
of its front becomes steep as a result of stretching caused by the cross-flow, causing the tip of the finger to become unstable and splits again, and the pattern repeats itself. Therefore, surface tension plays an essential dual role, it must be weak enough for the tip front to be unstable, but it is also the physical force causing the spreading and ensuing repeated branching \citep{tan_1988, carvalho_2013}.

@@

\medskip
Fluid flow dynamics through heterogeneous porous media are sensibly more complex than in homogeneous media and strongly depend on problem properties. Studies by \citet{langtangen_1992} on fluid flow dynamics in heterogeneous porous media (analytical and numerical solutions) demonstrated that hyperbolic Buckley-Leverett model is inherently unstable, \ie the expected uniform interface front collapses as any perturbation in physical parameters are imposed to the problem.

Spatial variation in geological formations occurs in all length-scales, where heterogeneity characteristics in small length-scales (\ie pore) are statistically embedded into permeability (absolute and relative) and porosity parameters. Such multi-scale heterogeneity induces preferential flow pathways and plays a significant role in the onset of fluid instabilities as it triggers fingers formation and their accelerated growth \citep[see][]{ewing_1989,tchelepi_1994}.

@@

%%%%%%%%%%%%%
\section{Conclusions}\label{Section:Conclusion}
% Short (1 paragraph) summary of the paper with focus on the main theoretical aspects that were proved by our work in the present manuscript;
The purpose of this paper is to numerically investigate the dynamics of viscous flow instabilities in porous media. Simulations involving formation and growth of viscous fingers were conducted with the CVFEM-based multi-fluid flow simulator Fluidity/IC-FERST model. The model is based upon a dual consistent pressure-velocity representation in CV and FEM spaces and a novel family of FE-pairs, \PN[n]{m} and \PNDG[n]{m}. Saturation and other scalar fields are discretised in CV space and may be fully discontinuous across finite element boundaries. %Material absolute permeability is piecewise constant in FE space to ensure the correct representation of the surface.
In order to strongly enforce discretised Darcy equations at the boundaries between elements and control volumes, a directional-weighted flux-limited was introduced to take into account discontinuities risen from both, numerical formulation (based on discontinuous-Galerkin finite element method, DGFEM) and control volumes within finite elements.

During immiscible multi-fluid flow displacements, the expected uniform front is perturbed if fluids have different densitites and/or viscositiesleading to an uneven front with elongations at the outside edge of the fluid interface

viscous fingering occurs when the viscosity ratio is greater than unity. As surface tension becomes weak, the interface is stressed and becomes unstable leading to the formation of fingers.
